---
title: Machine Learning Methods for Modeling and Classification of fashion MNIST
authors:
  - name: David Blumenstiel
    department: Department of Data Science and Information Systems
    affiliation: CUNY School of Professional Studies
    location: NY, NY 1001
    email: blumenstieldavid@gmail.com
  - name: Bonnie Cooper
    department: Department of Biological Sciences
    affiliation: SUNY College of Optometry
    location: NY, NY 10036
    email: bcooper@sunyopt.edu
  - name: Robert Welk
    department: Department of Data Science and Information Systems
    affiliation: CUNY School of Professional Studies
    location: NY, NY 1001
    email: robert.j.welk@gmail.com
  - name: Leo Yi
    department: Department of Data Science and Information Systems
    affiliation: CUNY School of Professional Studies
    location: NY, NY 1001
    email: leo.yi35@spsmail.cuny.edu  
abstract: |
  Fashion MNIST is a clothing classification dataset that is popular for deep learning and computer vision applications. In this report, we use Fashion MNIST to apply multiple machine learning methods reviewed in Data622: Machine Learning and Big Data, an elective course for the Masters of Science in Data Science at CUNY School of Professional Studies. We explore approaches to reduce the dimensionality of the data by engineering new descriptive features and performing Principal Components Analysis. We then follow this up with machine learning methods for classification such as Support Vector Machine and a Convolutional Neural Network. We find that 
keywords:
  - fashion MNIST
  - machine learning
  - classification
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
---

# Introduction

Fashion MNIST is a clothing classification dataset that builds in complexity in comparison to the classic MNIST dataset. MNIST is a dataset of handwritten digits that has been a go-to dataset for benchmarking various image processing and machine learning algorithms [@lecun1998gradient]. Classification algorithms applied to MNIST revolutionized the field of image processing in the 90s [@krizhevsky2009learning]. However, contemporary machine learning methods can achieve 97% accuracy. Convolutional neural nets score as high as 99.7% accurate. As a result, MNIST is now considered too easy and has also been used thoroughly. Fashion MNIST was developed as an alternative.

Fashion MNIST can serve as a direct drop-in replacement for MNIST. Fashion MNIST and MNIST are both labeled data that share the same dataset size: 60,000 training images and 10,000 test images. Additionally, Fashion MNIST and MNIST images share the same dimensions and structure: 10 distinct categories of grayscale images with 28x28 pixel size. Due to the complexity and variety of the images, Fashion MNIST is a more challenginf dataset for machine learning algorithms [@xiao2017fashion].

In your report, be sure to:

- describe the problem you are trying to solve.

- describe your datasets and what you did to prepare the data for analysis. 

- methodologies you used for analyzing the data

- why you did what you did

- make your conclusions from your analysis. Please be sure to address the business impact (it could be of any domain) of your solution.


# Headings: first level
\label{sec:headings}

You can use directly LaTeX command or Markdown text. 

LaTeX command can be used to reference other section. See Section \ref{sec:headings}.
However, you can also use **bookdown** extensions mechanism for this.

## Headings: second level

You can use equation in blocks

$$
\xi _{ij}(t)=P(x_{t}=i,x_{t+1}=j|y,v,w;\theta)= {\frac {\alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}{\sum _{i=1}^{N} \sum _{j=1}^{N} \alpha _{i}(t)a^{w_t}_{ij}\beta _{j}(t+1)b^{v_{t+1}}_{j}(y_{t+1})}}
$$

But also inline i.e $z=x+y$

### Headings: third level

Another paragraph. 


# Examples of citations, figures, tables, references
\label{sec:others}

You can insert references. Here is some text [@kour2014real; @kour2014fast] and see @hadash2018estimate.

The documentation for \verb+natbib+ may be found at

You can use custom blocks with LaTeX support from **rmarkdown** to create environment.

::: {.center latex=true}
  <http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}>
:::

Of note is the command \verb+\citet+, which produces citations
appropriate for use in inline text.  

You can insert LaTeX environment directly too.

\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}

produces

\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}


## Figures

You can insert figure using LaTeX directly. 

See Figure \ref{fig:fig1}. Here is how you add footnotes. [^Sample of the first footnote.]

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{4cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
  \label{fig:fig1}
\end{figure}

But you can also do that using R.

```{r fig2, fig.cap = "Another sample figure"}
plot(mtcars$mpg)
```

You can use **bookdown** to allow references for Tables and Figures.


## Tables

Below we can see how to use tables. 

See awesome Table~\ref{tab:table} which is written directly in LaTeX in source Rmd file.

\begin{table}
 \caption{Sample table title}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
  \label{tab:table}
\end{table}

You can also use R code for that.

```{r}
knitr::kable(head(mtcars), caption = "Head of mtcars table")
```


## Lists

- Item 1
- Item 2 
- Item 3
